\documentclass[a4paper,11pt,notitlepage]{article}

\input{preamble/all.tex}
 
\author{Karl D. Asmussen (\texttt{twk181})}
\date{\today}
\title{Multidimensional Arrays: Formally and Cleverly Sliced}
\addbibresource{refs.bib}

\begin{document}
\First{
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{abstract}
  We attempt to formalize the notion of a multidimensional array by Category
  theory, and investigate the semantics of array operations. We then
  implement an array library in Rust, and construct slices of multidimensional arrays
  using linear functions in homogenous coordinates using modules over the ring of integers.
\end{abstract}
\end{minipage}
\end{center}
\newpage
}

\section{Towards a Formalism}
\label{sec:to-form}

Several useful datastructures can be modeled as free functors, and in programming
languages like Haskell, it is quite explicit that they are. In particular, the List
type is explicitly both a Functor and a Monoid, since Haskell's lists are free monoids over
a set:

\begin{lstlisting}[style=Normal,language=Haskell]
data [] a = a : [a] | []

instance Functor [] where
  fmap f = map f

instance Monoid [] where
  mzero = []
  mappend = (++)
\end{lstlisting}\footnote{\lstinline[style=Normal,language=Haskell]{data [a] = a : [a] | []} is technically
not valid haskell. The list type has syntactic sugar to make it so.}

In essence, list is a monoid, with the empty list as identity and concatenation as multiplication, as well
as a functor taking functions between element types to the mapping operation.

My goal is to do something similar, but to mutlidimensional arrays. At time of writing, most multidimensional
array libraries either concern themselves exclusively with linear algebra, or use a semantics of operations that
concerns itself exclusively with utility --- not theoretical foundation.

\subsection{Shape of an Array}
\label{sub:shape}

Arrays are, in a very deep sense, functions. While we generally expect a function to be based on
a defining equation, set theory teaches us that functions can be arbitrary, and indeed even non-computable.
Arrays are (when used in computing) obviously computable, but not equationally so. An array of random numbers
taken from atmospheric noise cannot be generated by any feasible computation process.

Arrays like the common \verb`vector<T>` of C++, \verb`Vec<T>` of Rust, \verb`Array` of Ruby, \verb`list`
of Python, and so on are partial functions with the natural numbers
as domains (or alternately, total functions with a finite subset of the natural numbers as domains.)

(\textit{Aside:} indeed, hash maps are also functions, although with different domains. They are, however
built internally on arrays. More on that later.)



\Middle{\begin{center} End. \end{center}}

\Last{}
\end{document}
